---
title: "Assignment 1"
author: "Mel Miller"
date: "9/11/2020"
output: html_document
---

I first loaded the following libraries: I’ll be using the following libraries:

```{r, results = 'hide', message = FALSE}
library (ggplot2)
library (ggthemes)
library (scales)
library (tidycensus)
library (tidyverse)

person_vars_2018 <- pums_variables %>%
  distinct(year, survey, var_code, 
           var_label, data_type, level) %>%
  filter(level == "person", year == 2018, survey == "acs1")

```

For this assignment, I used a truncated dataset based on the dataset I created for Quant Assignment 1. I selected the following 6 variables, which are all filtered for older adults (age 65 and older) in Massachusetts: 

1. Age (AGEP): continuous  
2. Sex (SEX): categorical  
3. Race (RAC1P): categorical  
4. Housing tenure (TEN): categorical  
5. Total income, past 12 months (PINCP): continuous
6. Number of persons in respective household (NP): continuous  

```{r, include=FALSE}
person_data <- get_pums(variables = c("AGEP",
                                      "SEX", 
                                      "RAC1P",
                                      "TEN", 
                                      "PINCP",
                                      "NP"),
                        state = "MA", 
                        year = 2018, 
                        survey = "acs1",
                        recode = TRUE) %>%
  filter(AGEP > 64) %>%
  select(AGEP, SEX_label, RAC1P_label, TEN_label, PINCP, NP)
```

#1. Pie charts: race and housing tenure

I ran two pie charts to serve as an introduction to my dataset. The first is for race, and the second is for housing tenure. In both charts there are clear majorities: White for race, and owning for housing tenure (owning free and clear, followed by owning with mortgage or loan).

```{r}

pie_chart <- ggplot(person_data, aes(x = 1, fill = RAC1P_label)) +
  geom_bar(position = "fill") +
  coord_polar(theta = "y") +
  scale_y_continuous(name = "Number of people") + 
  scale_fill_brewer(name= "Race",
                    palette="Blues")+
  theme_void() 

pie_chart

pie_chart <- ggplot(person_data, aes(x = 1, fill = TEN_label)) +
  geom_bar(position = "fill") +
  coord_polar(theta = "y") +
  scale_y_continuous(name = "Number of people") +
  scale_fill_manual(name = "Housing Tenure",
                    values=c("#16a085", "#DAF7A6", "#FFC300", "#c70039", "#900c3f")) +
  theme_void() 
  
pie_chart

```
##Recoding values
After seeing the breakdown of race data in this dataset, I recoded this variable into: White, Black or African American, Asian, and Other. 

```{r}
person_data <- as.data.frame(person_data)

person_data$scode [person_data$RAC1P=="1"] <- "White"
person_data$scode [person_data$RAC1P=="2"] <- "Black"
person_data$scode [person_data$RAC1P=="6"] <- "Asian"
person_data$scode [person_data$RAC1P=="3"] <- "Other"
person_data$scode [person_data$RAC1P=="5"] <- "Other"
person_data$scode [person_data$RAC1P=="8"] <- "Other"
person_data$scode [person_data$RAC1P=="9"] <- "Other"

colnames(person_data)[colnames(person_data) == 'scode'] <- "Race"

```

##Pie chart: race recoded
The pie chart for the recoded race variable represents these four categories. Despite collapsing important categories, I hope this new variable will aid in the visualization of this and other data.
```{r}
pie_chart <- ggplot(person_data, aes(x = 1, fill = Race)) +
  geom_bar(position = "fill") +
  coord_polar(theta = "y") +
  scale_y_continuous(name = "Number of people") +
  scale_fill_brewer(name= "Race",
                    palette="Blues")+
  theme_void() 

pie_chart

```

#2. Scatterplot with trendline: Number of people in household by age
This chart started as a scatter plot with a trend line, but since the y-axis is a continuous variable with values from 1 to 13, the dots did not tell much of a story. I think it's interesting to see this chart without the dots, and look to the trend line to show how number of people in household varies by age. 
```{r} 
ggplot(person_data, 
       aes(x = NP,
           y = AGEP)) +
  stat_smooth(color = "black", linetype = 2, size = 0.5) +
  scale_x_continuous(name = "Number of people in household",
                     breaks = seq(1, 20, by = 1)) +
  scale_y_continuous(name = "Age") +
  theme_bw() 
```

## 3. Scatterplot with multiple trendlines: Number of people in household by age, by race
This scatter plot shows what I removed in the previous figure - the actual data points. This chart also incorporates race, and indicates that number of people in household decreases with age, with this difference most obvious for White folks. 

```{r}
ggplot(person_data, 
       aes(x = NP,
           y = AGEP,
           color = Race,
           fill = Race)) +
  geom_point() +
  stat_smooth(method = "lm") +
  scale_x_continuous(name = "Number of People in Household",
                     breaks = seq(0, 15, by = 1)) +
  scale_y_continuous(name = "Age",
                     breaks = seq(10, 100, by = 10)) +
  scale_color_discrete(name = "Race") +
  scale_fill_discrete(name = "Race") +
  theme_bw() 
```

#4. Box plot: Housing Tenure by Age
With this box plot, we see that the average age of folks who own their homes is lower than any other group. The oldest average age report "N/A" for housing tenure. 

```{r}
ggplot(person_data, 
       aes(x = TEN_label,
           y = AGEP,
           color = TEN_label)) +
  geom_boxplot() +
  scale_x_discrete(name = "Housing Tenure",
                    labels = c("N/A",
                               "Occupied without rent",
                               "Owned",
                               "Owned with mortgage",
                               "Rented")) +
  scale_y_continuous(name = "Age") +
  scale_fill_manual(values=c("#16a085", "#DAF7A6", "#FFC300", "#c70039", "#900c3f")) +
  theme_bw() +
  theme(legend.position = "none")

```
##5. Violin plot with scatter plot: 
Housing Tenure by number of people in household. Note: initially tried box plot but this enables a better understanding (although still complicated) 
```{r}
ggplot(person_data, 
       aes(x = TEN_label,
           y = NP,
           color = TEN_label)) +
  geom_violin() +
  geom_point(position = "jitter",
             alpha = 0.5,
             size = 0.5) +
  scale_x_discrete(name = "Housing Tenure",
                    labels = c("N/A",
                               "Occupied without rent",
                               "Owned",
                               "Owned with mortgage",
                               "Rented")) +
  scale_y_continuous(name = "Number of people in household",
                     breaks = seq(1, 15, by = 1)) +
  scale_fill_manual(values=c("#16a085", "#DAF7A6", "#FFC300", "#c70039", "#900c3f")) + 
  theme_bw() +
  theme(legend.position = "none")
```


#6. Violin plot: Housing Tenure by Income, using a logarithmic scale

```{r}
ggplot(person_data, 
       aes(x = TEN_label,
           y = PINCP,
           fill = TEN_label)) +
  geom_violin() +
  scale_x_discrete(name = "Housing Tenure",
                    labels = c("N/A",
                               "Occupied without rent",
                               "Owned",
                               "Owned with mortgage",
                               "Rented")) +
  scale_y_continuous(name = "Income",
                     trans = "log",
                     breaks = c(10, 100, 1000, 10000, 100000, 1000000),
                     labels = c("10", "100", "1,000", "10,000", "100,000", "1,000,000")) +
scale_fill_manual(values=c("#16a085", "#DAF7A6", "#FFC300", "#c70039", "#900c3f"))
  theme_bw() +
  theme(legend.position = "none")
```


#7. Scatterplot: Household Size by Income, Using a Logarithmic scale

```{r}
ggplot(person_data, 
       aes(x = NP,
           y = PINCP)) +
  geom_point() +
  stat_smooth(color = "black", linetype = 2, size = 0.5) +
  scale_x_continuous(name = "Household Size") +
  scale_y_continuous(name = "Income",
                     trans = "log",
                     breaks = c(0, 10, 100, 1000, 10000, 100000, 1000000),
                     labels = c("0", "10", "100", "1,000", "10,000", "100,000", "1,000,000"))+
  theme_bw() 
```

#8. Dot plot: Race and Number of people in household

```{r}
ggplot(person_data, 
       aes(x = Race,
           y = NP,
           color = Race)) +
  geom_point(position = "jitter",
             size = 0.5,
             alpha = 0.5) +
  scale_x_discrete(name = "Race") +
  scale_y_continuous((name = "Number of People in Household"),
                     breaks = seq(1, 20, by = 1)) +
  theme_bw() +
  theme(legend.position = "none")
```

#9. Stacked Bar chart to show race and housing tenure

```{r}
ggplot(person_data, aes(x = Race, fill = TEN_label)) +
  geom_bar() +
  scale_y_continuous(name = "Number of people") +
  scale_x_discrete(name = "Race") +
  scale_fill_discrete(name = element_blank()) +
  theme_bw() 
```

#10. Bar charts
##Bar cart with doge
Since there are so many white people, used "doge" function. I also switched from a linear to a log scale. 

```{r}
ggplot(person_data, aes(x = Race, fill = TEN_label)) +
  geom_bar(position = "dodge") +
  scale_y_continuous(name = "Number of people",
                     trans = "log",
                     breaks = c(30, 300, 3000)) +
  scale_x_discrete(name = "Race") +
  scale_fill_discrete(name = element_blank()) +
  theme_bw() 
```
  
##Bar chart without doge
But it's also interesting to see home tenure by race, to really see the discrepency in home ownership. 

```{r}
ggplot(person_data, aes(x = Race, fill = TEN_label)) +
  geom_bar(position = "fill") +
  scale_y_continuous(name = "Percent of home tenure",
                     breaks = seq(0, 1, by = 0.1),
                     labels = paste(seq(0,100, by = 10), "%", sep ="")) +
  scale_x_discrete(name = "Race") +
  scale_fill_discrete(name = element_blank()) +
  theme_bw() 
```




Print to PDF. once you do so, release all clipping masks. (object>clipping masks>release). now the graph is easier to edit. 
```{r}
pdf("name_file", width = 5, height = 5)
name_file

``` 
can also print to .png


can customize within GG themes
in code: #figure: commenting it out. computer won't run that line.