---
title: "Assignment 3"
author: "Mel Miller"
date: "9/23/2020"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: hide
---

# Loading the libraries and data

To create my dataset to use for this assignment, I'll first load the following libraries:

```{r, results = 'hide', message = FALSE}
library (sf)
library (tidyverse)
library (tidycensus)
library (ggthemes)
library (ggspatial)
library (units)
library (nngeo)
```

I'll read in five datasets from the Boston Open Data portal. I've decided to focus on open space and libraries in order to better understand the open space and community resources offered in the communities I'm focusing on for studio.  

Just as Carole did in the tutorial, I've set "quiet=TRUE" on each call of "st_read()" to allow for an easier read of this file.

```{r}
nhoods <- st_read("http://bostonopendata-boston.opendata.arcgis.com/datasets/3525b0ee6e6b427f9aab5d0a1d0a1a28_0.kml?outSR={%22latestWkid%22:2249,%22wkid%22:102686}", quiet = TRUE)

water <- st_read("http://bostonopendata-boston.opendata.arcgis.com/datasets/2b3c0fa13b1c468eb702a3645fcc6bcb_5.kml", quiet = TRUE)

openspace <- st_read ("http://bostonopendata-boston.opendata.arcgis.com/datasets/2868d370c55d4d458d4ae2224ef8cddd_7.kml", quiet = TRUE)

trees <- st_read ("http://bostonopendata-boston.opendata.arcgis.com/datasets/ce863d38db284efe83555caf8a832e2a_1.kml", quiet = TRUE)

libraries <- st_read ("http://bostonopendata-boston.opendata.arcgis.com/datasets/cb00f9248aa6404ab741071ca3806c0e_6.kml?outSR={%22latestWkid%22:2249,%22wkid%22:102686}", quiet = TRUE)
```

# Transforming the data

Since I'm interested in distance (and probably area calculations), I've followed Carole's lead and transformed my data to the Massachusetts State Plane, based on the NAD83 AEA.

```{r}
MA_state_plane <- "+proj=lcc +lat_1=41.71666666666667 +lat_2=42.68333333333333 +lat_0=41 +lon_0=-71.5 +x_0=200000 +y_0=750000 +ellps=GRS80 +units=m +no_defs"

nhoods <- nhoods %>%
  st_transform(MA_state_plane)

water <- water %>%
  st_transform(MA_state_plane)

openspace <- openspace %>%
  st_transform(MA_state_plane)

trees <- trees %>%
  st_transform(MA_state_plane)

libraries <- libraries %>%
  st_transform(MA_state_plane)
```

Here's a map of all this data:

```{r}
ggplot(water) +
  geom_sf(fill = "lightblue", color = NA) +
  geom_sf(data = trees, color = "darkgreen", size = 0.01) +
  geom_sf(data = openspace, color = "lightgreen") +
  geom_sf(data = libraries, color = "purple", size = 0.1) +
  geom_sf(data = nhoods, fill = NA, color = "gray") +
  theme_map () +
  annotation_scale()
```  

```{r}
ggplot(openspace) +
  geom_sf(fill = "lightgreen", color = NA) +
  geom_sf (data = water, fill = "lightblue", color = NA) +
  geom_sf(data = libraries, color = "purple", size = 0.1) +
  geom_sf(data = nhoods, fill = NA, color = "gray")
```  

# Creating a buffer

How many libraries in Boston are there within 50 meters of open space?

To figure this out, I'll create a new polygon layer representing a 50-meter buffer around the open space in Boston. 

```{r}
openspace_buffer <- st_buffer(openspace, dist = 50) %>%
  st_union()

ggplot(openspace_buffer) +
  geom_sf() +
  theme_map()
```

#Subsetting points with a polygon

Now I'll create a dataframe that only includes open spaces that are located within the library buffer. I'll draw them on top of the library buffer.

```{r}
libraries_openspace <- libraries[openspace_buffer,]
  
ggplot(openspace_buffer) +
  geom_sf() +
  geom_sf(data = libraries_openspace, 
          color = "darkmagenta") +
  theme_map()
```

Now I'll join the two dataframes for open space.

```{r}
libraries <- libraries %>%
  st_join(libraries_openspace) %>%
  mutate (by_openspace = !is.na(Name.y))
```

I'll calculate how many libraries are within 50 meters of open space:

```{r}
n_openspace_libraries <-sum(libraries$by_openspace)

n_openspace_libraries
```

10 libraries are within 50 meterse of open space. I'll now calculate what percent of all libraries in Boston this represents.

```{r}
n_libraries <- length (libraries$by_openspace)
pct_openspace_libraries <- n_openspace_libraries / n_libraries
pct_openspace_libraries
```

Nearly 40% of libraries are within 50 meters of open space. I'll combine all of this work onto one map.

```{r}
left_side  <- st_bbox(libraries)$xmin
top_side <- st_bbox(libraries)$ymax

ggplot(water) +
  geom_sf(fill = "lightblue", color = NA) +
  geom_sf(data = libraries, size = 1,
          aes(color = by_openspace)) +
  scale_color_manual(values = c("darkmagenta", "violet"),
          name = "Boston Libraries\nby distance to open space", 
          labels = c("No library within 50 m",
                     "Library within 50 m")) +
  annotation_scale(location = "br") +
  annotation_north_arrow(location = "tr",
                         style = north_arrow_minimal()) +
  annotate(geom = "text", x = left_side, 
           y = top_side, 
           label = paste("Of the ", 
                         prettyNum(n_libraries, big.mark = ","),
                         " libraries in Boston\n", 
                         prettyNum(n_openspace_libraries, big.mark = ","),
                         " (", 
                         prettyNum(100*pct_openspace_libraries, digits = 0),
                         "%) are within 50\nmeters of open space.",
                         sep = ""),
           hjust = 0, vjust = 0, size = 3) +
  theme_map() +
  theme(panel.background = element_rect(fill = "cornsilk1"),
        legend.background = element_rect(fill = alpha("white", 0.5), 
                                         color = "gray"))
```

Now, I'm interested in knowing: 1) How many libraries are in each Boston neighborhood, and 2) How library proximity to open space differs in each neighborhood.  

First, I'll explore this first question of how many libraries are in each Boston neighborhood.

**Note, since it's just 1-5, I think each # should be assigned a specific color.**
```{r}
nhoods <- nhoods %>%
  mutate(num_libraries = lengths(st_covers(nhoods, libraries)))

ggplot(nhoods) +
  geom_sf(color = NA, 
          aes(fill = num_libraries)) +
  scale_fill_viridis_c(name = "Boston neighborhoods\nby number of libraries",
                       breaks = breaks <- seq(1, 5, by = 1),
                       labels = paste(prettyNum(breaks, big.mark = ","),
                                      "libraries")) +
  annotation_scale(location = "br") +
  annotation_north_arrow(location = "tr",
                         style = north_arrow_minimal()) +
theme_map() +
  theme(legend.background = element_rect(fill = alpha("white", 0.5), 
                                         color = "gray"))
```

# Calculating areas and densities 

What is the density of libraries (the number of libraries per square km)?

*Not sure why this isn't working!!*

```{r}

nhoods <- nhoods %>%
  mutate(area = set_units(st_area(nhoods), km^2)) %>%
  mutate(lib_dens = as.numeric(num_libraries / area))

ggplot(nhoods) +
  geom_sf(color = NA, 
          aes(fill = lib_dens)) +
    scale_fill_viridis_c(name = 
                           "Boston neighborhoods\nby library density",
                         breaks = breaks <- seq(0, 2, by = .5),
                       labels = paste(prettyNum(breaks, big.mark = ","),
                                      "libraries per square km")) +
  annotation_scale(location = "br") +
  annotation_north_arrow(location = "tr",
                         style = north_arrow_minimal()) +
theme_map() +
  theme(legend.position = "right",
    legend.background = element_rect(fill = alpha("white", 0.5), 
                                         color = "gray"))
```

# Finding the closest point

How far is each library from open space? 

*left off here. not working*
```{r}
libraries <- libraries %>%
  mutate(openspace_dist = st_nn(libraries, openspace, 
                           returnDist = TRUE)$dist) %>%
  mutate(openspace_dist = as.numeric(openspace_dist))
```

Calculating the average distance from a library to the nearest open space.

```{r}
avg_openspace_dist <- mean (libraries$openspace_dist)

avg_openspace_dist
```

A map to see how far each library is from open space.

```{r}
right_side <- st_bbox(libraries)$xmax
left_side  <- st_bbox(libraries)$xmin
top_side <- st_bbox(libraries)$ymax
bottom_side <- st_bbox(libraries)$ymin


ggplot(water) +
  geom_sf(fill = "lightblue", color = NA) +
  geom_sf(data = libraries, size = 1,
          aes(color = openspace_dist)) +
  coord_sf(xlim = c(left_side, right_side), 
           ylim = c(bottom_side, top_side), expand = FALSE) +
  scale_color_viridis_c(name = 
                          "Boston Library meters\nby distance to openspace") +
  annotation_scale(location = "br") +
  annotation_north_arrow(location = "tr",
                         style = north_arrow_minimal()) +
  annotate(geom = "text", x = left_side + 300, 
           y = top_side - 550, 
           label = paste("On average, a Boston library\nis ", 
                         prettyNum(avg_openspace_dist, digits = 3),
                         " meters from open space.",
                         sep = ""),
           hjust = 0, vjust = 0, size = 3) +
  theme_map() +
  theme(panel.background = element_rect(fill = "cornsilk1"),
        legend.background = element_rect(fill = alpha("white", 0.5), 
                                         color = "gray"))
```

# Identifying overlapping polygons

Which neighborhoods have open space in them?

```{r}
nhoods <- nhoods %>%
  mutate(num_openspace = lengths(st_overlaps(nhoods, openspace))) %>%
  mutate(has_openspace = num_openspace > 0)

```

How many neighborhoods overlap with openspace?

```{r}
n_openspace_nhoods <- sum(nhoods$has_openspace)

n_openspace_nhoods
```

Mapping this data.

```{r}
left_side  <- st_bbox(openspace)$xmin
top_side <- st_bbox(openspace)$ymax

ggplot(openspace) +
  geom_sf(fill = "lightgreen", color = NA) +
  geom_sf(data = nhoods,
          aes(fill = has_openspace)) +
  scale_fill_manual(values = c("cornsilk1", "darkseagreen1"),
          name = "Boston Neighborhoods\nby presence of open space", 
          labels = c("Neighborhood without\nopenspace",
                     "Neighborhood with\nopenspace")) +
  annotation_scale(location = "br") +
  annotation_north_arrow(location = "tr",
                         style = north_arrow_minimal()) +
  annotate(geom = "text", x = left_side, 
           y = top_side - 1000, 
           label = paste(n_openspace_nhoods ,
                         "of Boston's", 
                         length(nhoods$Name),
                         "neighborhoods contain\nor overlap with", 
                         "openspae."),
           hjust = 0, vjust = 0, size = 3) +
  theme_map() +
  theme(panel.background = element_rect(fill = "gray"),
        legend.background = element_rect(fill = alpha("white", 0.5), 
                                         color = "gray"))
```