---
title: "Assignment 1"
author: "Mel Miller"
date: "9/15/2020"
output: html_document
---

# Preparing the Dataset

To create my dataset to use for this assignment, I first loaded the following libraries:

```{r, results = 'hide', message = FALSE}
library (ggplot2)
library (ggthemes)
library (tidycensus)
library (tidyverse)

person_vars_2018 <- pums_variables %>%
  distinct(year, survey, var_code, 
           var_label, data_type, level) %>%
  filter(level == "person", year == 2018, survey == "acs1")

```

For this assignment, I used a truncated dataset based on the one I created for Quant Assignment 1. I selected the following 6 variables, which are all filtered for older adults (age 65 and older) in Massachusetts: 

1. Age (AGEP)  
2. Sex (SEX)  
3. Race (RAC1P)  
4. Housing tenure (TEN)  
5. Total income, past 12 months (PINCP)
6. Household size (NP)

```{r, include=FALSE}
person_data <- get_pums(variables = c("AGEP",
                                      "SEX", 
                                      "RAC1P",
                                      "RAC1P",
                                      "TEN", 
                                      "PINCP",
                                      "NP"),
                        state = "MA", 
                        year = 2018, 
                        survey = "acs1",
                        recode = TRUE) %>%
  filter(AGEP > 64) %>%
  select(AGEP, SEX_label, RAC1P_label, RAC1P, TEN_label, PINCP, NP)
```

The first 10 rows of my data are shown below. You'll notice the variable for race (RAC1P) is included twice in the code below. As you'll see shortly, I included this twice so that I would be able to recode it into a new variable.
```{r}
person_data
```
# Visualizing the Data
Using this dataset, I created 13 charts. These charts are categorized into the ten sections that follow. 

## 1. Pie charts: Race and housing tenure

I ran two pie charts to serve as an introduction to my dataset. The first is for race, and the second is for housing tenure. In both charts there are clear majorities: White for race, and owning for housing tenure (owning free and clear, followed by owning with mortgage or loan). For the second race group, the full label is as follows: "American Indian and Alaska Native tribes specified; or American Indian or Alaska Native, not specified and no other races."

```{r}

pie_chart <- ggplot(person_data, aes(x = 1, fill = RAC1P_label)) +
  geom_bar(position = "fill") +
  coord_polar(theta = "y") +
  scale_y_continuous(name = "Number of people") + 
  scale_fill_brewer(name= "Race",
                    labels = c("American Indian",
                               "American Indian and Alaska Native",
                               "Asian",
                               "Black or African American",
                               "Some Other Race",
                               "Two or more races",
                               "White alone"),
                    palette="Dark2")+
  theme_void() 

pie_chart

pie_chart <- ggplot(person_data, aes(x = 1, fill = TEN_label)) +
  geom_bar(position = "fill") +
  coord_polar(theta = "y") +
  scale_y_continuous(name = "Number of people") +
  scale_fill_brewer (name = "Housing tenure",
                     labels = c("N/A",
                               "Occupied without rent",
                               "Owned",
                               "Owned with mortgage",
                               "Rented"),
                    palette="Set2") +
  theme_void() 
  
pie_chart

```
  
### Recoding values
After seeing the breakdown of race data in this dataset, I recoded this variable into: White, Black or African American, Asian, and Other. Even though this recoding collapses important categories, I believe simplifying the data in this way helps with this and other visualizations that use this variable.

```{r}
person_data <- as.data.frame(person_data)

person_data$scode [person_data$RAC1P=="1"] <- "White"
person_data$scode [person_data$RAC1P=="2"] <- "Black"
person_data$scode [person_data$RAC1P=="6"] <- "Asian"
person_data$scode [person_data$RAC1P=="3"] <- "Other"
person_data$scode [person_data$RAC1P=="5"] <- "Other"
person_data$scode [person_data$RAC1P=="8"] <- "Other"
person_data$scode [person_data$RAC1P=="9"] <- "Other"

colnames(person_data)[colnames(person_data) == 'scode'] <- "Race"

```

### Pie chart: Race recoded
The pie chart for the recoded race variable represents the previously mentioned four categories. 

```{r}

pie_chart <- ggplot(person_data, aes(x = 1, fill = Race)) +
  geom_bar(position = "fill") +
  coord_polar(theta = "y") +
  scale_y_continuous(name = "Number of people") +
  scale_fill_brewer(name = "Race",
                    palette="Dark2") +
  theme_void() 

pie_chart

```

## 2. Scatter plot with trendline: Household size by age
This chart started as a scatter plot with a trendline, but since the y-axis is a continuous variable with values from 1 to 13, the dots did not tell much of a story. I think it is interesting to see this chart without the dots, and look to the trend line to show how household size varies by age. Overall, this chart shows that the household size decreases with age, and that this relationship has more variation as the number of people in the household increases. I believe this is because of the fact that there are many fewer people who live with more than 6 people in their homes, which is when we begin to see the spread of the data substantially increase.
```{r} 
ggplot(person_data, 
       aes(x = NP,
           y = AGEP)) +
  stat_smooth(color = "black", linetype = 2, size = 0.5) +
  scale_x_continuous(name = "Household size",
                     breaks = seq(1, 20, by = 1)) +
  scale_y_continuous(name = "Age",
                     breaks = seq(60, 100, by = 5)) +
  theme_tufte() 
```

## 3. Scatter plot with multiple trendlines: Household size by race
This scatter plot differs by the previous figure in several ways. First, it shows the actual point data, which I had removed from the previous figure. This chart also incorporates race, enabling viewers to conceptualize the relationship with household size, age, and race. As with the previous graph, here we see that household size decreases with age. We also see this relationship in terms of race, with what I believe the clearest take away is that the number of people in household decreases with age the most for White people.
```{r}
ggplot(person_data, 
       aes(x = NP,
           y = AGEP,
           color = Race,
           fill = Race)) +
  geom_point() +
  stat_smooth(method = "lm") +
  scale_x_continuous(name = "Household size",
                     breaks = seq(0, 15, by = 1)) +
  scale_y_continuous(name = "Age",
                     breaks = seq(60, 100, by = 10)) +
  scale_fill_brewer(name = "Race",
                    palette="Dark2") +
  scale_color_brewer(name = "Race",
                    palette="Dark2") +
  theme_bw() 
```

## 4. Box plot: Housing tenure by age
I believe this box plot provides the best summary of the data so far. Here we see that, on average, those who own their home with a mortgage are younger than any other housing tenure group, while those who indicated "N/A" for their housing tenure are the oldest. While I was initially not sure what this "N/A" group represented, this chart makes me think it may be individuals who live in assisted living communities who, even though they usually pay rent, this housing tenure is different than the typical conception of "renting."

```{r}
ggplot(person_data, 
       aes(x = TEN_label,
           y = AGEP,
           color = TEN_label)) +
  geom_boxplot() +
  scale_x_discrete(name = "Housing tenure",
                    labels = c("N/A",
                               "Occupied without rent",
                               "Owned",
                               "Owned with mortgage",
                               "Rented")) +
  scale_y_continuous(name = "Age",
                     breaks = seq(60, 100, by = 5)) +
  scale_color_brewer(palette="Set2") +
  theme_bw() +
  theme(legend.position = "none")

```
  
## 5. Violin plot with scatter plot: Housing tenure by household size
While this figure has the same issue mentioned earlier with the other scatter plots, I think its helpful to since it shows how many people exist within each category of housing tenure and household size. Even though this high concentration at points does not allow for an easy understanding of the shape of the violin plot, I  think it still reveals important takeaways. These include: the largest households are owned with mortgage, and all N/A housing tenures contain just one person. I believe this further supports my thought that these individuals are living in assisted living facilities.

```{r}
ggplot(person_data, 
       aes(x = TEN_label,
           y = NP,
           color = TEN_label)) +
  geom_violin() +
  geom_point(position = "jitter",
             alpha = 0.5,
             size = 0.5) +
  scale_x_discrete(name = "Housing tenure",
                    labels = c("N/A",
                               "Occupied without rent",
                               "Owned",
                               "Owned with mortgage",
                               "Rented")) +
  scale_y_continuous(name = "Number of people in household",
                     breaks = seq(1, 15, by = 1)) +
  scale_color_brewer (palette = "Set2") + 
  theme_bw() +
  theme(legend.position = "none")
```


## 6. Violin plot: Housing tenure by income, using a logarithmic scale
This graph surprises me in how little variation it shows, although perhaps this is because of the logarithmic scale. There are differences, though, including the fact that the N/A housing tenure group has the least variation in income, and home owners (both those who own their homes and those who have mortgages) have the greatest variation in income.


```{r}
ggplot(person_data, 
       aes(x = TEN_label,
           y = PINCP,
           fill = TEN_label)) +
  geom_violin() +
  scale_x_discrete(name = "Housing tenure",
                    labels = c("N/A",
                               "Occupied without rent",
                               "Owned",
                               "Owned with mortgage",
                               "Rented")) +
  scale_y_continuous(name = "Income",
                     trans = "log",
                     breaks = c(10, 100, 1000, 10000, 100000, 1000000),
                     labels = c("10", "100", "1,000", "10,000", "100,000", "1,000,000")) +
  scale_fill_brewer(palette="Set2") +
  theme_bw() +
  theme(legend.position = "none")
```


## 7. Scatter plot: Household size by income, using a logarithmic scale

Again, this chart has a similar issue as previous charts that use scatter plots for household data, That being said, I believe it shows that income variation is the greatest for smaller households, and that, for the most part, incomes decrease as household size increases.

```{r}
ggplot(person_data, 
       aes(x = NP,
           y = PINCP)) +
  geom_point() +
  stat_smooth(color = "black", linetype = 2, size = 0.5) +
  scale_x_continuous(name = "Household size",
                     breaks = seq(1, 20, by = 1)) +
  scale_y_continuous(name = "Income",
                     trans = "log",
                     breaks = c(0, 10, 100, 1000, 10000, 100000, 1000000),
                     labels = c("0", "10", "100", "1,000", "10,000", "100,000", "1,000,000"))+
  theme_tufte() 
```

## 8. Dot plot: Race and household size

I think this chart is misleading since there are so many more white people in the sample than any other race and, at a quick glance, it appears to show that White people have the smallest households (in terms of number of people). While I do not think it allows for a fair understanding of the relationship between race and household size, it does allow for insight into household sizes within race. For example, we see that 1- and 2-person households are the most common across all race groups, and that the range in household size is smallest among the Other race category.

```{r}

ggplot(person_data, 
       aes(x = Race,
           y = NP,
           color = Race)) +
  geom_point(position = "jitter",
             size = 0.5,
             alpha = 0.5) +
  scale_x_discrete(name = "Race") +
  scale_y_continuous((name = "Household size"),
                     breaks = seq(1, 20, by = 1)) +
  scale_color_brewer(palette="Dark2") +
  theme_bw() +
  theme(legend.position = "none")
```

## 9. Stacked bar chart: Race and housing tenure

This chart is also limited in its visualization due to the large number of White folks in the sample. While its hard to parse out the differences in housing tenure among Asian, Black, and Other races, we can clearly see the differences in housing tenure for White people.

```{r}
ggplot(person_data, aes(x = Race, fill = TEN_label)) +
  geom_bar() +
  scale_y_continuous(name = "Number of people",
                     breaks = c(0, 2500, 5000, 7500, 10000, 12500),
                     labels = c("0", "2,500", "5,000", "7,500", "10,000", "12,500")) +
  scale_x_discrete(name = "Race") +
  scale_fill_brewer(name="Housing tenure",
                    labels = c("N/A",
                               "Occupied without rent",
                               "Owned",
                               "Owned with mortgage",
                               "Rented"), 
                    palette = "Set2")+
  theme_bw() 
```

## 10. Bar charts: Race and housing tenure
### Bar cart with dodge
Since there are so many White people, I used the "dodge" function to create what I think of as a more traditional bar chart. I also created this chart using a linear scale to provide a better understanding of differences in housing tenure by race. This chart allows for a clear understanding of not only how housing tenure differs by each race, but also how housing tenure differs within each race.

```{r}
ggplot(person_data, aes(x = Race, fill = TEN_label)) +
  geom_bar(position = "dodge") +
  scale_y_continuous(name = "Number of people",
                     trans = "log",
                     breaks = c(30, 300, 3000),
                     labels= c("30", "300", "3,000")) +
  scale_x_discrete(name = "Race") +
  scale_fill_brewer(name = "Housing tenure",
                    labels = c("N/A",
                               "Occupied without rent",
                               "Owned",
                               "Owned with mortgage",
                               "Rented"), 
                    palette="Set2")+
  theme_bw() 
```
  
### Bar chart without dodge, utilizing fill
In thinking about how housing tenure differs within each race, I decided to create one last bar chart. I think this bar chart combines the best features of the previous two charts to allow for an understanding of housing tenure by race. Here we see that nearly 50% of folks in the Other race group are renters, compared to 10% of white people. Another finding this chart shows is that the percentage of folks who occupy their home without rent is small across all race groups.

```{r}
ggplot(person_data, aes(x = Race, fill = TEN_label)) +
  geom_bar(position = "fill") +
  scale_y_continuous(name = "Percent of housing tenure",
                     breaks = seq(0, 1, by = 0.1),
                     labels = paste(seq(0,100, by = 10), "%", sep ="")) +
  scale_x_discrete(name = "Race") +
  scale_fill_brewer (name = "Housing tenure",
                    labels = c("N/A",
                               "Occupied without rent",
                               "Owned",
                               "Owned with mortgage",
                               "Rented"), 
                     palette="Set2")+
  theme_bw() 
```