---
title: "Assignment 4"
author: "Mel Miller"
date: "10/5/2020"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: hide
---

# Loading the libraries and data

I'll first load all relevant libraries.

```{r, warning = FALSE, message = FALSE}
library (osmdata)
library (opentripplanner)
library (tidyverse)
library (sf)
library (ggthemes)
library (ggspatial)
```

## Load locations

I'll load in the locations for Boston public schools, and then filter:

1. For only public schools in Dorchester
2. For only public elementary (K-3) schools in Dorchester

```{r, results = 'hide', message = FALSE}

BPS_Geojson <- st_read ("http://bostonopendata-boston.opendata.arcgis.com/datasets/1d9509a8b2fd485d9ad471ba2fdb1f90_0.geojson?outSR=%7B%22latestWkid%22%3A2249%2C%22wkid%22%3A102686%7D")

Dorchester_BPS <- BPS_Geojson %>%
  filter (CITY == "Dorchester")

Dorchester_EL_BPS <- BPS_Geojson %>%
  filter (CITY == "Dorchester", SCH_TYPE == "ES")

```


## Get street data

I'll get street data for Dorchester, Boston, MA.

```{r, warning = FALSE, message = FALSE}
MA_state_plane <- "+proj=lcc +lat_1=41.71666666666667 +lat_2=42.68333333333333 +lat_0=41 +lon_0=-71.5 +x_0=200000 +y_0=750000 +ellps=GRS80 +units=m +no_defs"

dot_street_query <- opq(bbox = 'Dorchester MA USA')%>%
  add_osm_feature (key = 'highway')

dot_street_query %>%
  osmdata_xml (file = 'OTP/Graphs/Default/dot_streets.osm')

dot_street_features <- dot_street_query %>%
  osmdata_sf()

dot_streets <- dot_street_features$osm_lines %>%
  st_transform(crs=MA_state_plane)

```

### Plot Dorchester streets

```{r}
ggplot(dot_streets) + 
  geom_sf () +
  theme_map ()
```

### Set up OpenTripPlanner

```{r, warning = FALSE, results = 'hide', message = FALSE}
path_otp <- otp_dl_jar("OTP")
```

```{r, results = 'hide'}
path_data <-file.path (getwd(), "OTP")
path_otp <- paste (path_data, "otp.jar", sep ="/")

otp_build_graph(otp = path_otp, dir = path_data, memory = 1024) 
```

### Launch OpenTripPlanner

```{r, message=FALSE, results='hide'}
otp_setup(otp = path_otp, dir = path_data, memory =1024)
```

### Connect to server

```{r}
otpcon <- otp_connect()
```

# Create isochrones

I'll first create isochrones for areas within a five-minute bike and five-minute walk from all Boston Public Schools in Dorchester.

```{r}
iso_5min_walk <-
    otp_isochrone(otpcon = otpcon, fromPlace = Dorchester_BPS, 
                mode = "WALK", cutoffSec = 300) %>%
  st_transform(crs = MA_state_plane) %>%
  mutate(mode = "walk")

iso_5min_bike <- 
  otp_isochrone(otpcon = otpcon, fromPlace = Dorchester_BPS, 
                mode = "BICYCLE", cutoffSec = 300) %>%
  st_transform(crs = MA_state_plane) %>%
  mutate(mode = "bike")

iso_all_modes <- rbind(iso_5min_bike, iso_5min_walk)
```

```{r}
right_side <- st_bbox(iso_all_modes)$xmax
left_side  <- st_bbox(iso_all_modes)$xmin
top_side <- st_bbox(iso_all_modes)$ymax
bottom_side <- st_bbox(iso_all_modes)$ymin

ggplot(iso_all_modes) +
  annotation_map_tile(zoomin = 0, progress = "none") +
  geom_sf(aes(fill = mode), alpha = 0.5) +
  geom_sf(data = Dorchester_BPS) +
  coord_sf(xlim = c(left_side, right_side), 
           ylim = c(bottom_side, top_side), expand = FALSE) +
  scale_fill_viridis_d(name = "Area that is reachable within 5 minutes",
                       labels = c("by bike", "by foot")) +
  theme_map() +
  labs(caption = "Basemap Copyright OpenStreetMap contributors") +
  theme(legend.position= "right")
```

From this point forward I'll focus on just elementary schools (K-3) in Dorchester. Note that I've chosen to not include K-8 and K-12 schools in Dorchester.

```{r}
iso_5min_walk <-
    otp_isochrone(otpcon = otpcon, fromPlace = Dorchester_EL_BPS, 
                mode = "WALK", cutoffSec = 300) %>%
  st_transform(crs = MA_state_plane) %>%
  mutate(mode = "walk")

iso_5min_bike <- 
  otp_isochrone(otpcon = otpcon, fromPlace = Dorchester_EL_BPS, 
                mode = "BICYCLE", cutoffSec = 300) %>%
  st_transform(crs = MA_state_plane) %>%
  mutate(mode = "bike")

iso_all_modes_EL <- rbind(iso_5min_bike, iso_5min_walk)

otp_stop()
```

```{r}
right_side <- st_bbox(iso_all_modes_EL)$xmax
left_side  <- st_bbox(iso_all_modes_EL)$xmin
top_side <- st_bbox(iso_all_modes_EL)$ymax
bottom_side <- st_bbox(iso_all_modes_EL)$ymin

ggplot(iso_all_modes_EL) +
  annotation_map_tile(zoomin = 0, progress = "none") +
  geom_sf(aes(fill = mode), alpha = 0.5) +
  geom_sf(data = Dorchester_EL_BPS) +
  coord_sf(xlim = c(left_side, right_side), 
           ylim = c(bottom_side, top_side), expand = FALSE) +
  scale_fill_viridis_d(name = "Area that is reachable within 5 minutes",
                       labels = c("by bike", "by foot")) +
  theme_map() +
  labs(caption = "Basemap Copyright OpenStreetMap contributors") +
  theme(legend.position= "right")
```

## Thunderforestlandscape basemap
While I think this map is too busy, I think it's interesting to see the isochrones contextualized with the major transit stops and parks/open space.

```{r}
right_side <- st_bbox(iso_all_modes_EL)$xmax
left_side  <- st_bbox(iso_all_modes_EL)$xmin
top_side <- st_bbox(iso_all_modes_EL)$ymax
bottom_side <- st_bbox(iso_all_modes_EL)$ymin

ggplot(iso_all_modes_EL) +
  annotation_map_tile(zoomin = 0, type="thunderforestlandscape", progress = "none") +
  geom_sf(aes(fill = mode), alpha = 0.5) +
  geom_sf(data = Dorchester_EL_BPS) +
  coord_sf(xlim = c(left_side, right_side), 
           ylim = c(bottom_side, top_side), expand = FALSE) +
  scale_fill_discrete(name = "Area that is reachable within 5 minutes",
                       labels = c("by bike", "by foot"),
                       type = c("white", "black")) +
  theme_map() +
  labs(caption = "Basemap Copyright OpenStreetMap contributors") +
  theme(legend.position= "right")
```

## Cartolight basemap

This map provides an understanding of major roads and neighborhood squares (e.g., Fields Corner). I think this map does the best job so far at helping visualize where K-3 elementary schools are, along with their accessibility (by bike and foot).

```{r}
right_side <- st_bbox(iso_all_modes_EL)$xmax
left_side  <- st_bbox(iso_all_modes_EL)$xmin
top_side <- st_bbox(iso_all_modes_EL)$ymax
bottom_side <- st_bbox(iso_all_modes_EL)$ymin

ggplot(iso_all_modes_EL) +
  annotation_map_tile(zoomin = 0, type="cartolight", progress = "none") +
  geom_sf(aes(fill = mode), alpha = 0.5) +
  geom_sf(data = Dorchester_EL_BPS) +
  coord_sf(xlim = c(left_side, right_side), 
           ylim = c(bottom_side, top_side), expand = FALSE) +
  scale_fill_viridis_d(name = "Area that is reachable within 5 minutes",
                       labels = c("by bike", "by foot")) +
  theme_map() +
  labs(caption = "Basemap Copyright OpenStreetMap contributors") +
  theme(legend.position= "right")
```

## Hillshade basemap

This last map uses shading to show hills, or the lack of, in Dorchester.

```{r}
right_side <- st_bbox(iso_all_modes_EL)$xmax
left_side  <- st_bbox(iso_all_modes_EL)$xmin
top_side <- st_bbox(iso_all_modes_EL)$ymax
bottom_side <- st_bbox(iso_all_modes_EL)$ymin

ggplot(iso_all_modes_EL) +
  annotation_map_tile(zoomin = 0, type="hillshade", progress = "none") +
  geom_sf(aes(fill = mode), alpha = 0.5) +
  geom_sf(data = Dorchester_EL_BPS) +
  coord_sf(xlim = c(left_side, right_side), 
           ylim = c(bottom_side, top_side), expand = FALSE) +
  scale_fill_viridis_d(name = "Area that is reachable within 5 minutes",
                       labels = c("by bike", "by foot")) +
  theme_map() +
  labs(caption = "Basemap Copyright OpenStreetMap contributors") +
  theme(legend.position= "right")
```

# Calculate and compare isochrone areas

I'll calculate the area of the biking and walking isochrones to visualize the relationship between the two.

```{r}
iso_areas <- iso_all_modes_EL %>%
  mutate(area = st_area(iso_all_modes_EL)) %>%
  st_set_geometry(NULL) %>%
  pivot_wider(names_from = mode, values_from = area) 

ggplot(iso_areas, 
       aes(x = as.numeric(walk), y = as.numeric(bike))) +
  geom_point() +
  scale_x_continuous(name = 
            "Area within a five-minute walking distance\nof a public elementary (K-3) school\n(square km)",
            breaks = breaks <- seq(10000, 100000, by = 10000),
            labels = breaks / 1000000) +
  scale_y_continuous(name = 
            "Area within a five-minute biking distance\nof a public elementary (K-3) school\n(square km)",
            breaks = breaks <- seq(0, 800000, by = 100000),
            labels = breaks / 1000000) +
  theme_bw()
```
